---
title: "ASL Temp change calcs"
author: "Abby Lewis"
date: "2022-08-29"
output: html_document
---

This file calculates volume-weighted hypolimnetic oxygen demand during the stratified period at each lake. 

Table of contents:
Step 1: Load data and packages
Step 2: Calculate the rate of change in volume-weighted oxygen concentrations during the stratified period at each lake (VHOD)
Step 3: Calculate temperature-corrected oxygen demand values
Step 4: Calculate trends in oxygen demand over time


Step 1: Load data and packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(rLakeAnalyzer)
library(openair)

#Load data and stratification boundaries. This file is created by "03 - Stratified avgs.Rmd"
full_with_thermo <- read.csv("../Compiled data/Stratified_period_data_with_thermo.csv")%>%
  mutate(Date=as.Date(Date))
#Load lake metadata
lat_long <- read.csv("https://pasta-s.lternet.edu/package/data/eml/edi/1029/9/fadd3eaa25b5fdd1fc4efba70e660579")
```


Step 2: Calculate the rate of change in hypo temps during the stratified period at each lake (hypo_warming)
```{r}
#Total n to start: 609
length(unique(full_with_thermo$LakeID))

#Calculate mean hypo temps
hypo_temp <- full_with_thermo%>%
  filter(Depth_m>hypo_depth)%>%
  group_by(Date, LakeID)%>%
  dplyr::summarize(Temp_C = mean(Temp_C, na.rm = T), #mean across all hypolimnetic depths
                   )%>%
  mutate(Year= year(Date))%>%
  group_by(LakeID, Year)%>%
  filter(!is.na(Temp_C))%>%
  ungroup()

length(unique(hypo_temp$LakeID)) #603 lakes left

#Calculate the rate of change in hypo temps using lm
hypo_warming <- hypo_temp%>%
  mutate(Year = year(Date))%>%
  group_by(Year, LakeID)%>%
  dplyr::summarize(n = n(),
            warming_rate = lm(Temp_C~Date)$coefficients[2],
            p = summary(lm(Temp_C~Date))$coefficients[8],
            r2 = summary(lm(Temp_C~Date))$r.squared)%>%
  filter(n>=3) %>% #Need at least 3 points
  filter(!is.na(warming_rate))
length(unique(hypo_warming$LakeID[!is.na(hypo_warming$warming_rate)])) #502

#How many rows get removed by filtering so R2>0.5? 246
removed_by_r2 <- hypo_warming%>%
  filter(r2>.5)
nrow(removed_by_r2[!is.na(removed_by_r2$warming_rate),]) #6497
nrow(hypo_warming[!is.na(hypo_warming$warming_rate),]) #7899
7899-6497
(7899-6497)/7899

#Finish QAQC
good_warming <- removed_by_r2
length(unique(good_warming$LakeID)) #492

#Success!
write.csv(good_warming, "../Compiled data/Hypo_warming_rate.csv", row.names = F)
```

Step 4: Calculate trends in warming rate over time. This is not currently part of the manuscript, but I thought the analysis was interesting
```{r}
split_year = 1800

#Function
sen_slope_custom <- function(df,var){
  output = df%>%
    mutate(era = ifelse(Year<split_year,"before","after"))%>%
    group_by(LakeID,era)%>%
    dplyr::summarize(n = n(),
              trend = NA,
              sig = NA,
              min_year = NA,
              max_year = NA)
  for(lake in unique(df$LakeID)){
    for(era in c("before","after")){
      print(paste(lake,era))
      filt = df%>%
        filter(LakeID==lake)%>%
        mutate(date = as.POSIXct(paste0(Year,"-01-01")))
      if(era == "before"){
        filt = filt%>%
          ungroup()%>%
          filter(Year<split_year)%>%
          dplyr::select(-Year)
      }else{
        filt = filt%>%
          ungroup()%>%
          filter(Year>split_year)%>%
          dplyr::select(-Year)
      }
      if(length(unique(year(filt$date)))>=10){#Only calculate a trend if there are 10 years of data
        sen = TheilSen(filt, pollutant = var, avg.time = "year", plot = F)$data$res2
        output$trend[output$LakeID==lake&output$era == era]<-sen$slope[1]
        output$sig[output$LakeID==lake&output$era == era]<-sen$p[1]
        output$min_year[output$LakeID==lake&output$era == era]<-min(year(filt$date))
        output$max_year[output$LakeID==lake&output$era == era]<-max(year(filt$date))
      }
    }
  }
  return(output)
}

#Calculate trends
vw_demand_sen <- sen_slope_custom(good_warming,"warming_rate")

#Plot
vw_demand_sen%>%
  filter(!is.na(trend),
         sig<0.05)%>%
  mutate(type = ifelse(trend<0,"Decreasing","Increasing"))%>%
  ggplot(aes(x = type))+
  geom_bar()+
  xlab("")+
  ggtitle("Of lakes with significant demand trends:")

vw_demand_sen%>%
  ungroup()%>%
  filter(!is.na(trend))%>%
  mutate(n_tot = n())%>%
  filter(sig<0.05)%>%
  mutate(type = ifelse(trend<0,"Decreasing","Increasing"))%>%
  group_by(type)%>%
  dplyr::summarize(n = n(),
                   pct = n/unique(n_tot)*100)

```
